Title Text

### 2015-03-12 docs/paper.tex
A bit about incremental solvers

@@

\begin{Verbatim}
(declare-fun x () Int)
(assert (= 3 x) 8)
(assert (not (= x 5)))
(check-sat)
\end{Verbatim}

@@

@@

The second bullet is useful because it helps us avoid inferring type
schemes with unsatisfiable constraints.  Suppose, for example, that
we had just a single wanted constraint: \Verb"(x + 5) ~ 2".  Since we
are working with natural numbers, this constraint has no solution, so
we will not be able to solve it. As a result, we may end up inferring
a type like this:
\begin{Verbatim}
someFun :: forall x. (x + 5) ~ 2 => ...
\end{Verbatim}
While, technically, this is not wrong, it is undesirable because the type
error is clearly in the definition of \Verb"someFun", but we would delay
reporting the error until the function is called.  So, it is better if
we noticed that the wanted constraints are not satisfiable, are report
and error immediately.

The importance of the third bullet is that it enables collaboration between
the SMT solver plug-in, and the rest of GHC (i.e., the main constraint solver,
and also other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd have to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  This equality will then be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3",
which in turn can be discharged by the custom solver for the \Verb"KnownNat"
class.  This ``collaboration'' between different solvers is quite common
in GHC.
We discuss this observation further in Section~\ref{modular-typechecker}.

>>>>

\subsection{The Core Functionality}

@@

\begin{Verbatim}
(declare-fun x () Int)
(assert (= (+ 3 x) 8))
(assert (not (= x 5)))
(check-sat)
\end{Verbatim}

@@

\subsection{Incremental Solvers}

Many SMT solvers have support for asserting formulas {\em incrementally}.
This means that the solver performs a little work every time a new formula
is asserted, rather than collecting all formulas and doing all the work
in batch mode, once we ask about satisfiability.  Furthermore, incremental
solvers have the ability to mark a particular solver state, and then revert
back to it, using a stack discipline.  Here is an example:

\begin{Verbatim}
(declare-fun x () Int)
(assert (= (+ 5 x) 3))
(push 1)
  (assert (>= x 0))
  (check-sat)
(pop 1)
(check-sat)
\end{Verbatim}

This example asks the solver two questions, and the answer to the first
one is \Verb"unsatisfiable", while the answer to the second one is
\Verb"satisfiable".

The \Verb"push" command instructs to solver to save its state, then
proceed as normal.  When the solver encounters a \Verb"pop" command,
it reverts to the last saved state.

This functionality is extremely useful when we want to ask many questions
that are largely the same, and differ only in a few assertions:
we can perform the majority of the work once, and then use
\Verb"push" and \Verb"pop" to just assert the differences.
In our small example, the work for the first assertion shared in
both calls to \Verb"check-sat".

@@

The second bullet is useful because it helps us avoid inferring type
schemes with unsatisfiable constraints.  Suppose, for example, that
we had just a single wanted constraint: \Verb"(x + 5) ~ 2".  Since we
are working with natural numbers, this constraint has no solution, so
we will not be able to solve it. As a result, we may end up inferring
a type like this:
\begin{Verbatim}
someFun :: forall x. (x + 5) ~ 2 => ...
\end{Verbatim}
While, technically, this is not wrong, it is undesirable because the type
error is clearly in the definition of \Verb"someFun", but we would delay
reporting the error until the function is called.  So, it is better if
we noticed that the wanted constraints are not satisfiable, are report
an error immediately.

The importance of the third bullet is that it enables collaboration between
the plug-in and the rest of GHC (i.e., the main constraint solver,
and other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd have to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  This equality will then be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3",
which in turn can be discharged by the custom solver for the \Verb"KnownNat"
class.  Such collaboration between different solvers is quite common.
Interestingly, this looks a lot like the collaboration between decision
procedures in an SMT solver!  We discuss this observation further
in Section~\ref{modular-typechecker}.
